Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2018-04-18 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
update

\begin{abstract} 
%CAUSE --> EFFECT --> MOTIVATION (P1)
  A critical aspect of multiphase flow in porous media is the displacement efficiency that measures the amount of fluid that can be pushed by another fluid driven by pressure gradient. Migration of contaminants and reservoir waterflooding are typical applications where understanding the dynamics of immiscible fluid displacement helps mitigating water resources contamination and improving hydrocarbons production, respectively. Due to large viscosity ratios, flow instabilities at fluids' interface may arise leading to the formation of fingers -- uneven front with elongation at the outside edge of fluids interface, with strong impact on the displacement efficiency.  
%CAUSE --> EFFECT --> GAPS IN KNOWLEDGE --> AIMS OF THE WORK (P2)
Initial studies of viscous instabilities in Hele-Shaw cells indicated that the development of fingers mostly depends on mobility ratio and capillary number, however heterogeneity of the porous domain may also affect the onset of instabilities. Therefore, the main aim of this work is to numerically investigate formation and growth of viscous fingers in heterogeneous porous media. 
%CAUSE --> EFFECT --> FEW TECHS USED  (P3)
The model used here is based on a novel control volume finite element method (CVFEM) formulation with a family of FE-pairs, \PN[n]{m} specially tailored for Darcean flows. Dynamic mesh adaptivity enables capturing fingers development whilst saving computational overheads. Numerical experiments were performed to investigate the impact of viscosity ratio and heterogeneity on Saffmann-Taylor instabilities.
%CAUSE --> EFFECT -->  MAIN FINDINGS  (P4)
Numerical simulations demonstrated that the heterogeneity of the domain triggers the early-onset formation of fingers under prescribed viscosity ratio conditions. Also, effective numerical capture of growth (in particular, tip-splitting) and coalescence of dendritic finger branching induced by large viscosity ratio largely depends on mesh resolution at the fluids interface.

@@

\medskip 
In this work, a novel CVFEM formulation, previously introduced by \citet{gomes_2017} (see also \citet{jackson_2015,salinas2015}), is used to numerically investigate formation and growth of viscous fingers in heterogeneous porous media. The continuity equation is embedded into the pressure equation to enforce mass conservation whilst ensuring that the force balance is preserved. A hybrid family of \PN[n]{m} triangular and tetrahedral FE pairs is used to discretise velocity and pressure. %, in which velocity is represented by $n$ th-order polynomials that are discontinuous across elements whereas pressure is represented by $m$ th-order polynomials that are continuous across the elements.
A sketch of the \PN[1]{2} FE-pair is shown in Fig.~\ref{fig:fem_cv}, in which velocity is represented by discontinuous and piecewise linear basis functions whereas pressure is interpolated through continuous and piecewise quadratic basis functions. Scalar fields are stored in CV space (Fig.~\ref{fig:fem_elem}). The dual pressure and velocity fields are represented simultaneously (through non-linear projections) in FE and CV spaces.

@@

\begin{eqnarray}
\left(\displaystyle\frac{\mu_{\alpha}S_{\alpha}}{{\mathbf K}\mathcal{K}_{r\alpha}}\right) {\mathbf u}_{\alpha} = \underline{\underline{\sigma}}_{\alpha} {\mathbf u}_{\alpha} = -\nabla p + \mathcal{S}_{u,\alpha}\hspace{1cm} \text{ with } x_{i}\in\Omega,\; t>0, \label{eqn:darcy_eqn} \\
\phi\displaystyle\frac{\partial S_{\alpha} }{\partial t} +   \nabla \cdot \left( {\mathbf u}_{\alpha}  S_{\alpha}\right) =  \mathcal{S}_{cty,\alpha}\hspace{1cm} \text{ with } x_{i}\in\Omega,\; t>0,\label{eqn:saturation_eqn}
\end{eqnarray}

\noindent respectively, where $\mu$, ${\bf K}$, $p$ and $\phi$ are viscosity, absolute permeability, pressure and porosity, respectively. ${\mathbf u}_{\alpha}$ is the saturation-weighted Darcy velocity of the $\alpha$-phase and $\mathcal{K}_{r,\alpha}$ is the relative permeability. $\mathcal{S}$ is the source term associated with the Darcy and continuity equations. $S_{\alpha}$ represents the saturation of the $\alpha$-phase with mass conservation constraints of $\sum\limits_{\alpha=1}^{\mathcal{N}_{p}} S_{\alpha} = 1$, where $\mathcal{N}_{p}$ denotes the number of phases. $\underline{\underline{\sigma}}_{\alpha}$ is an absorption-like term that represents the implicit linearisation of the viscous frictional forces.
%\begin{displaymath}
%\sum\limits_{\alpha=1}^{\mathcal{N}_{p}} S_{\alpha} = 1, 
%\end{displaymath}
%where $\mathcal{N}_{p}$ denotes the number of phases.

>>>>

\begin{abstract} 
%CAUSE --> EFFECT --> MOTIVATION (P1)
  A critical aspect of multiphase flow in porous media is the displacement efficiency that measures the amount of fluid that can be pushed by another fluid driven by pressure gradient. Migration of contaminants and reservoir waterflooding are typical applications where understanding the dynamics of immiscible fluid displacement helps mitigating water resources contamination and improving hydrocarbons production, respectively. Due to large viscosity ratios, flow instabilities at fluids' interface may arise leading to the formation of fingers, uneven front with elongation at the outside edge of fluids interface, with strong impact on the displacement efficiency.  
%CAUSE --> EFFECT --> GAPS IN KNOWLEDGE --> AIMS OF THE WORK (P2)
Initial studies of viscous instabilities in Hele-Shaw cells indicated that the development of fingers mostly depends on mobility ratio and capillary number, however heterogeneity of the porous domain may also affect the onset of instabilities. Therefore, the main aim of this work is to numerically investigate formation and growth of viscous fingers in heterogeneous porous media. 
%CAUSE --> EFFECT --> FEW TECHS USED  (P3)
The model used here is based on a novel control volume finite element method (CVFEM) formulation with a family of FE-pairs, \PN[n]{m} specially tailored for Darcean flows. Dynamic mesh adaptivity enables capturing fingers development whilst saving computational overheads. Numerical experiments were performed to investigate the impact of viscosity ratio and heterogeneity on Saffmann-Taylor instabilities.
%CAUSE --> EFFECT -->  MAIN FINDINGS  (P4)
Numerical simulations demonstrated that the heterogeneity of the domain triggers the early-onset formation of fingers under prescribed viscosity ratio conditions. Also, effective numerical capture of growth (in particular, tip-splitting) and coalescence of dendritic finger branching induced by large viscosity ratio largely depends on mesh resolution at the fluids interface.

@@

\medskip 
In this work, a novel CVFEM formulation, previously introduced by \citet{gomes_2017} (see also \citet{jackson_2015,salinas2015}), is used to numerically investigate formation and growth of viscous fingers in heterogeneous porous media. The continuity equation is embedded into the pressure equation to enforce mass conservation whilst ensuring that the force balance is preserved. A hybrid family of \PN[n]{m} triangular and tetrahedral FE pairs is used to discretise velocity and pressure. %, in which velocity is represented by $n$ th-order polynomials that are discontinuous across elements whereas pressure is represented by $m$ th-order polynomials that are continuous across the elements.
A sketch of the \PN[1]{2} FE-pair is shown in Fig.~\ref{fig:fem_cv}, in which velocity is represented by discontinuous and piecewise linear basis functions whereas pressure is interpolated through continuous and piecewise quadratic basis functions. Scalar fields are stored in CV space (Fig.~\ref{fig:fem_elem}). %The dual pressure and velocity fields are represented simultaneously (through non-linear projections) in FE and CV spaces.

@@

\begin{eqnarray}
\left(\displaystyle\frac{\mu_{\alpha}S_{\alpha}}{{\mathbf K}\mathcal{K}_{r\alpha}}\right) {\mathbf u}_{\alpha} = \underline{\underline{\sigma}}_{\alpha} {\mathbf u}_{\alpha} = -\nabla p + \mathcal{S}_{u,\alpha}\hspace{1cm} \text{ with } x_{i}\in\Omega,\; t>0, \label{eqn:darcy_eqn} \\
\phi\displaystyle\frac{\partial S_{\alpha} }{\partial t} +   \nabla \cdot \left( {\mathbf u}_{\alpha}  S_{\alpha}\right) =  \mathcal{S}_{u,\alpha}\hspace{1cm} \text{ with } x_{i}\in\Omega,\; t>0,\label{eqn:saturation_eqn}
\end{eqnarray}

\noindent respectively, where $\mu$, ${\bf K}$, $p$ and $\phi$ are viscosity, absolute permeability, pressure and porosity, respectively. ${\mathbf u}_{\alpha}$ is the saturation-weighted Darcy velocity of the $\alpha$-phase and $\mathcal{K}_{r,\alpha}$ is the relative permeability. $\mathcal{S_{ua})}$ is the source term associated with the Darcy and continuity equations. $S_{\alpha}$ represents the saturation of the $\alpha$-phase with mass conservation constraints of $\sum\limits_{\alpha=1}^{\mathcal{N}_{p}} S_{\alpha} = 1$, where $\mathcal{N}_{p}$ denotes the number of phases. $\underline{\underline{\sigma}}_{\alpha}$ is an absorption-like term that represents the implicit linearisation of the viscous frictional forces.
%\begin{displaymath}
%\sum\limits_{\alpha=1}^{\mathcal{N}_{p}} S_{\alpha} = 1, 
%\end{displaymath}
%where $\mathcal{N}_{p}$ denotes the number of phases.
